---  
title:   "Supplementary material 2: Statistical analysis."  
author:  "T. Leon, J. Domingo, A.L. Riffo-Campos and G. Ayala"  
date:  "`r   Sys.Date()`"  
address: Departamento de Estadística e I.O. Universidad de Valencia  
format:   
   pdf:  
    toc: true
    number-sections: true
    colorlinks: true  
references:
  - id: Oliveira2025
    type: article-journal
    title: "High-definition spatial transcriptomic profiling of immune cell populations in colorectal cancer"
    author:
      - family: Oliveira
        given: Michelli Faria de
      - family: Romero
        given: Juan Pablo
      - family: Chung
        given: Meii
      - family: Williams
        given: Stephen R.
      - family: Gottscho
        given: Andrew D.
      - family: Gupta
        given: Anushka
      - family: Pilipauskas
        given: Susan E.
      - family: Mohabbat
        given: Seayar
      - family: Raman
        given: Nandhini
      - family: Sukovich
        given: David J.
      - family: Patterson
        given: David M.
      - family: de Oliveira
        given: Michelli Faria
      - family: Taylor
        given: Sarah E. B.
    issued:
      year: 2025
      month: 6
      day: 1
    container-title: "Nature Genetics"
    volume: "57"
    number: "6"
    page: "1512–1523"
    ISSN: "1546-1718"
    DOI: "10.1038/s41588-025-02193-3"
    URL: "https://doi.org/10.1038/s41588-025-02193-3"
    
  - id: baddeley2005spatstat
    type: article-journal
    title: "spatstat: An R Package for Analyzing Spatial Point Patterns"
    author:
      - family: Baddeley
        given: A.
      - family: Turner
        given: R.
    container-title: "Journal of Statistical Software"
    volume: "12"
    number: "6"
    page: "1–42"
    issued:
      year: 2005
    DOI: "10.18637/jss.v012.i06"
    URL: "https://doi.org/10.18637/jss.v012.i06"


---  


# Packages and data sets

## Loading the required packages  

```{r}
#| label: supmat1

pacman::p_load(spatstat,igraph,combinat,parallel)
source("functions_srt.R")

```

## Creating the point pattern  

### Reading transcription event coordinates

The analysis is performed on different subimages delineated by an expert.
Specifically, we focused on the same expert-defined tumor microenvironment
regions previously analyzed in @Oliveira2025 in their Figure 7, where Xenium
in situ profiling was used to confirm the existence and spatial localization
of macrophage subtypes and clonally expanded T cells.
Images come from three patients (1, 2, and 5): areas c and e for patient 1,
areas d and f for patient 2, and area h for patient 5.

The first step is to convert the transcription event files generated as
described in Supplementary Material 1 into local spatial point patterns
handled by R (format .rda).

```{r}  
#| label: supmat2  
#| eval: false

for(i in c(1,2,5)){
    switch(as.character(i), 
           "1" = {areasn = c(0:4,"c","e")},
           "2" = {areasn = c(0:2,"d","f")},
           "5" = {areasn = c(0:2,"h")}
           ) 
    for(j in areasn){
        cat("Case ",i," Area ",j,"\n")
        dirbase = paste0("~/Nextcloud/SPATIAL_TRANSCRIPTOMICS/MARKED_RESULTS_AREAS_C",
                         i,"/C",i,"_Area",j,"/")
        x = do.ppp(dirbase=dirbase,case=i,ff=1,min.n = 10)
        save(x,file=paste0("../data/local/x_l_case_",i,"_area_",j,"_ff_",1,
                           ".rda"))
    }
}

```


# Statistical Analysis 

## Data-driven detection of atypical spatial interactions 

1.  For each pair of genes $(i,j)$, three interaction statistics are estimated:
the cross-type $K$-function ($K_{ij}(r)$), the cross-type $L$-function ($L_{ij}(r)$), and 
the cross-type pair correlation function ($g_{ij}(r)$).

2.  For each interaction statistic, a comparison envelope for the pair $(i,j)$ is constructed
from the corresponding functions of all other gene pairs, excluding $(i,j)$ itself.

3. If the observed statistic for the pair $(i,j)$ partially falls outside the comparison 
envelope, the interaction is \emph{flagged as atypical}. Values above the envelope indicate 
spatial attraction or aggregation, values below indicate spatial repulsion or inhibition, 
and values within the envelope suggest no meaningful departure from the typical patterns 
observed across the system.

4. These atypical interactions are then used to construct gene--gene graphs, in which the 
width of each edge is set proportional to the number of distance intervals for which the 
corresponding interaction lies outside its comparison envelope.

5. The spatial distances at which the observed interaction statistics fall outside the 
envelopes are also identified.

## Estimating cross functions  
Estimates of the cross-type summary functions are obtained from the observed point patterns
using the Kcross, Lcross, and pcfcross functions from the R package spatstat
(@baddeley2005spatstat), and the results are saved.

```{r}
#| label: supmat3
#| eval: false

detectCores()
rvalues = seq(0,152,2)
ff = 1
for(i in c(1,2,5)){
    switch(as.character(i),
           "1" = {areasn = c(0:4,"c","e")},
           "2" = {areasn = c(0:2,"d","f")},
           "5" = {areasn = c(0:2,"h")}
           )
    for(j in areasn){
        cat("Case ",i," Area ",j,"\n")
        file_x = paste0("../data/local/x_l_case_",i,"_area_",j,"_ff_",1,".rda")
        load(file_x)
        sel = which(!is.na(x))
        pairs1 = t(combn(sel,2))
        x1 = x[sel]
        X = do.call(superimpose, x1)
        pairs11 =  t(combn(length(sel),2))
        kc.l  = mclapply(1:nrow(pairs11),function(k){
           ## cat("k ",k,"\n")
            Kcross(X,i=levels(marks(X))[pairs11[k,1]],
                   j=levels(marks(X))[pairs11[k,2]],r=rvalues,
                   correction="border")
        })
        Lc.l  = mclapply(1:nrow(pairs11),function(k){
            ## cat("k ",k,"\n")
            Lcross(X,i=levels(marks(X))[pairs11[k,1]],
                   j=levels(marks(X))[pairs11[k,2]],r=rvalues,
                   correction="border")
        })
        pcf.l = mclapply(1:nrow(pairs11),function(k){
            ## cat("k ",k,"\n")
            pcfcross(X,i=levels(marks(X))[pairs11[k,1]],
                     j=levels(marks(X))[pairs11[k,2]],r=rvalues,
                     correction="iso")
        })
        kc = matrix(unlist(lapply(kc.l,function(x) x$border)),
                    ncol=length(rvalues),byrow=TRUE)
        Lc = matrix(unlist(lapply(Lc.l,function(x) x$border)),
                    ncol=length(rvalues),byrow=TRUE)
        pcf = matrix(unlist(lapply(pcf.l,function(x) x$iso)),
                     ncol=length(rvalues),byrow=TRUE)
        save(kc,file=paste0("results/kc_l_",i,"_ff_",ff,"_area_",j,
                            ".rda"))
        save(Lc,file=paste0("results/Lc_l_",i,"_ff_",ff,"_area_",j,
                            ".rda"))
        save(pcf,file=paste0("results/pcf_l_",i,"_ff_",ff,"_area_",j,
                             ".rda"))
    }
}

```

## Finding atypical spatial interactions

For each interaction statistic, a reference envelope for the pair $(i,j)$  is generated 
using the corresponding summary functions from all other gene pairs, excluding the pair $(i,j)$.  

If the observed statistic for $(i,j)$ extends partially beyond this reference envelope, 
the interaction is marked as atypical.  

Additionally, we record on which side of the envelope the statistic exits (above or below),
as this directionality provides further information about the nature of the deviation.  

```{r}
#| label: supmat4
#| eval: false

ff = 1
for(desc in c("kc","Lc","pcf")){
    for(i in c(1,2,5)){
        switch(as.character(i),
               "1" = {areasn = c(0:4,"c","e")},
               "2" = {areasn = c(0:2,"d","f")},
               "5" = {areasn = c(0:2,"h")}
               )
       for(j in areasn){
            cat("Case ",i," Area ",j,"\n")
            file_desc = paste0("results/local/",desc,"_l_",i,"_ff_",ff,"_area_",j,".rda")
            load(file_desc)
             switch(desc,
               "kc" = {sig = oneRow(x=kc,method="all")},
               "Lc" = {sig = oneRow(x=Lc,method="all")},
               "pcf" = {sig = oneRow(x=pcf,method="all")}
               )
            file_sig = paste0("results/local/sig_",desc,"_l_",i,"_ff_",ff,"_area_",j,".rda")
            save(sig,file=file_sig)
        }
    }
}



## side  
ff = 1
for(desc in c("pcf")){
    for(i in c(1,2,5)){
        switch(as.character(i),
               "1" = {areasn = c(0:4,"c","e")},
               "2" = {areasn = c(0:2,"d","f")},
               "5" = {areasn = c(0:2,"h")}
               )
        for(j in areasn){
            cat("Case ",i," fraction ", ff," area ",j,"\n")
            file_desc = paste0("results/local/",desc,"_l_",i,"_ff_",ff,"_area_",j,".rda")
            file_sig = paste0("results/local/sig_",desc,"_l_",i,"_ff_",ff,"_area_",j,".rda")
            load(file_desc)
            load(file_sig)
            switch(desc,
                   "kc" = {side = oneRowSide(x=kc,sig=sig)},
                   "Lc" = {side = oneRowSide(x=Lc,sig=sig)},
                   "pcf" = {side = oneRowSide(x=pcf,sig=sig)}
                   )
            file_sig_side = paste0("results/local/sig_side_",desc,"_l_",i,"_ff_",ff,"_area_",j,".rda")
            save(side,file=file_sig_side)
        }
    }
}
 

```


## Construction of graphs from genes with atypical interactions


### Mapping ENSEMBL IDs to gene symbols

First, we establish the correspondence between Ensembl identifiers and gene symbols to be used in the graphs.

```{r}
#| label: supmat5
#| eval: false

genes = read.csv("GSE280314_gene_panel.csv",header=TRUE,sep="\t")
ensembl = genes$id[1:100]
hgnc_symbol  = genes$name[1:100]
save(ensembl,hgnc_symbol,file="annotation.rda")

``` 

### Graphs with local interactions  

Next, we generate the graphs for local point patterns. 

```{r}
#| label: supmat6
#| eval: false

source("functions_srt.R")

ff = 1
for(desc in c("kc","Lc")){
    for(i in c(1,2,5)){
        switch(as.character(i), 
               "1" = {areasn = c(0:4,"c","e")},
               "2" = {areasn = c(0:2,"d","f")},
               "5" = {areasn = c(0:2,"h")}
               ) 
        for(j in areasn){
            cat("desc ",desc,"Case ",i," Area ",j,"\n")
            file_x = paste0("../data/local/x_l_case_",i,"_area_",j,"_ff_",1,".rda")
            file_sig = paste0("results/local/sig_",desc,"_l_",i,"_ff_",ff,"_area_",j,
                              ".rda")
            file_sig_side = paste0("results/local/sig_side_",desc,"_l_",i,"_ff_",ff,
                                   "_area_",j,".rda")
            file_annotation = "annotation.rda"
            g = do_graph(file_x,file_sig,file_sig_side,file_annotation)
            file_plot=paste0("figures/local/graph_",desc,"_l_",i,"_ff_",ff,"_area_",j,
                             ".png")
            png(file_plot)
            plot(g)
            dev.off()
            file_graph = paste0("figures/local/graph_",desc,"_l_",i,"_ff_",ff,"_area_",j,
                            ".graph")
            save(g,file=file_graph)
            metag = data.frame(V1 = E(g)$label1,V2 =  E(g)$label2,width = E(g)$width,
                                   out = factor(E(g)$out,levels=1:2,labels=c("low","up")))
            file_metag = paste0("figures/local/metagraph_",desc,"_l_",i,"_ff_",ff,"_area_",j,
                                    ".csv")
            write.csv(metag,file=file_metag)
        }
    }
}


```


## Envelopes  

### Intervals where the descriptive is out the envelopes  


```{r}
#| label: supmat7
#| eval: false

ff = 1
file_annotation ="annotation.rda"
load(file_annotation)
df_all = NULL
for(desc in c("kc","Lc")){
    for(i in c(1,2,5)){
        switch(as.character(i), 
               "1" = {areasn = c(0:4,"c","e")},
               "2" = {areasn = c(0:2,"d","f")},
               "5" = {areasn = c(0:2,"h")}
               ) 
        for(j in areasn){
            cat("desc ",desc,"Case ",i," Area ",j,"\n")
            file_x = paste0("../data/local/x_l_case_",i,"_area_",j,"_ff_",1,".rda")
            load(file_x)
            sel = which(!is.na(x))
            pairs1 = t(combn(sel,2))
            file_sig = paste0("results/local/sig_",desc,"_l_",i,"_ff_",ff,"_area_",j,
                              ".rda")
            load(file_sig)
            pairs1 = pairs1[sig,]
            sel1 = sort(unique(c(pairs1[,1],pairs1[,2])))
            file_sig_side = paste0("results/local/sig_side_",desc,"_l_",i,"_ff_",ff,
                                   "_area_",j,".rda")
            load(file_sig_side)
            file_range =paste0("results/local/ranges_",desc,"_l_",i,"_ff_",ff,"_area_",j,
                             ".csv")
            df = data.frame(desc= rep(desc,nrow(pairs1)),
                            case = rep(i,nrow(pairs1)),
                            window = rep(j,nrow(pairs1)),
                            gene1 = hgnc_symbol[pairs1[,1]],
                            gene2 = hgnc_symbol[pairs1[,2]],
                            lower = apply(side,1,function(u) min(which(u !=0))),
                            upper = apply(side,1,function(u) max(which(u !=0))))
            df_all = rbind(df_all,df)
            save(df,file = file_range)
        }
    }
}
save(df_all,file = "results/local/df_all.rda")

```

## Envelopes  

```{r}
#| label: supmat8
#| eval: false
load("results/local/df_all.rda")
df_all
```


### Local interactions  

```{r}
#| label: supmat9
#| eval: false
ff = 1

i=1;j="c";ff=1;desc="kc";genes=c("CXCL13","CXCL10");dir_figures ="figures/local/";
           file_annotation="annotation.rda"

p = do_env(i=1,j="c",ff=1,desc="kc",genes=c("CXCL13","CXCL10"),dir_figures ="figures/local/",
           file_annotation="annotation.rda")

do_env(i=1,j="e",ff=1,desc="kc",genes=c("SPP1","CHI3L1"),dir_figures ="figures/local/",
       file_annotation="annotation.rda")

do_env(i=1,j="e",ff=1,desc="kc",genes=c("SPP1","MMP9"),dir_figures ="figures/local/",
       file_annotation="annotation.rda")

do_env(i=1,j="e",ff=1,desc="kc",genes=c("MZB1","MUC5B"),dir_figures ="figures/local/",
       file_annotation="annotation.rda")

do_env(i=1,j="e",ff=1,desc="kc",genes=c("SPP1","MMP3"),dir_figures ="figures/local/",
       file_annotation="annotation.rda")

do_env(i=2,j="d",ff=1,desc="kc",genes=c("CHI3L1","REG3A"),dir_figures ="figures/local/",
       file_annotation="annotation.rda")

do_env(i=2,j="d",ff=1,desc="kc",genes=c("SPP1","COMP"),dir_figures ="figures/local/",
       file_annotation="annotation.rda")

do_env(i=2,j="f",ff=1,desc="kc",genes=c("CHI3L1","MMP12"),dir_figures ="figures/local/",
       file_annotation="annotation.rda")

do_env(i=2,j="f",ff=1,desc="kc",genes=c("CHI3L1","COMP"),dir_figures ="figures/local/",
       file_annotation="annotation.rda")

do_env(i=5,j="h",ff=1,desc="kc",genes=c("CXCL9","IDO1"),dir_figures ="figures/local/",
       file_annotation="annotation.rda")

do_env(i=5,j="h",ff=1,desc="kc",genes=c("CXCL9","GREM1"),dir_figures ="figures/local/",
       file_annotation="annotation.rda")


``` 
